---
title: "MidTerm_Telecom"
author: "Emma Horton"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

MidTerm Telecom Customer Churn Analysis
Emma Horton – Data Visualization Midterm Project
Repository: https://github.com/HumaDec/DataVizMidterm

For this project, I selected the Telco Customer Churn dataset (originally provided by IBM and available on Kaggle), which includes data on 7,043 telecom customers across 21 columns. Each row represents a customer, with information spanning:

Demographics (e.g., gender, senior citizen status)

Subscription details (e.g., contract type, internet service)

Billing metrics (e.g., monthly and total charges)

Churn status

This dataset was chosen because of its rich mix of categorical (e.g., Contract, Churn) and quantitative variables (e.g., Tenure, MonthlyCharges), which are well-suited for visualization and exploratory analysis.



Libraries Used: 
```{r libraries, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(plotly)
library(lsr)
library(broom)
library(htmlwidgets)

```

Data Loading
To ensure proper parsing of blank fields (especially in TotalCharges), I used na.strings = c(""). This helps convert empty strings into NA values.
```{r}
telco <- read.csv("Telco-Customer-Churn.csv", na.strings = c(""))

#str(telco)
#summary(telco)
#head(telco, 5)
```
Data Cleaning
Convert SeniorCitizen to a labeled factor

Convert multiple character variables to factors

Convert TotalCharges to numeric (after removing empty string values)

Standardize "No internet service"/"No phone service" to "No"

Drop customerID, which is not useful

```{r}
# SeniorCitizen to factor
telco$SeniorCitizen <- factor(telco$SeniorCitizen, levels=c(0,1), labels=c("No", "Yes"))

# Character columns to factors
cat_cols <- c("gender","Partner","Dependents","PhoneService",
              "MultipleLines","InternetService","OnlineSecurity","OnlineBackup",
              "DeviceProtection","TechSupport","StreamingTV","StreamingMovies",
              "Contract","PaperlessBilling","PaymentMethod","Churn")

telco[cat_cols] <- lapply(telco[cat_cols], factor)

# Fix TotalCharges: convert to numeric
telco$TotalCharges <- as.numeric(as.character(telco$TotalCharges))

# Replace "No internet service" and "No phone service" with "No", replace "No phone service" with "No"
no_internet_cols <- c("OnlineSecurity","OnlineBackup","DeviceProtection",
                      "TechSupport","StreamingTV","StreamingMovies")
for(col in no_internet_cols) {
    telco[[col]] <- factor(ifelse(telco[[col]]=="No internet service", "No", 
                                  as.character(telco[[col]])))
}
telco$MultipleLines <- factor(ifelse(telco$MultipleLines=="No phone service", "No", 
                                     as.character(telco$MultipleLines)))
# Drop customerID
telco$customerID <- NULL

# Verify results
#sum(is.na(telco$TotalCharges))  
#levels(telco$MultipleLines)        
#levels(telco$OnlineSecurity)       

```
Set Color Pallet
To maintain consistency and improve the visual appeal of plots, I defined a custom color palette for key variable levels
```{r}
project_colors <- c(
  "Yes" = "#E74C3C",    
  "No" = "#3498DB",         
  "Female" = "#9B59B6",    
  "Male" = "#1ABC9C",     
  "Month-to-month" = "#F39C12",  
  "One year" = "#2ECC71",       
  "Two year" = "#34495E",       
  "DSL" = "#16A085",        
  "Fiber optic" = "#D35400",
  "No" = "#95A5A6"          
)

```

Exploratory Analysis
First, I examined how contract type relates to customer churn. This mosaic plot visualizes the relationship between contract type and churn status. The plot and chi-square test reveal a significant relationship: Month-to-month customers churn at a much higher rate than those with longer-term contracts.

```{r}
# Cross-tab Churn by Contract
table(telco$Churn, telco$Contract)

# Chi-square
chisq_test <- chisq.test(table(telco$Churn, telco$Contract))
chisq_test

# mosaic 
mosaic <- mosaicplot(table(telco$Contract, telco$Churn),
           color = c(project_colors["No"], project_colors["Yes"]),
           main = "Churn vs Contract Type",
           xlab = "Contract Type", ylab = "Churn")


png("mosaic.png", width = 800, height = 600)
mosaicplot(table(telco$Contract, telco$Churn),
           color = c(project_colors["No"], project_colors["Yes"]),
           main = "Churn vs Contract Type",
           xlab = "Contract Type", ylab = "Churn")
dev.off()

```

To better understand customer retention, I explored the distribution of tenure (i.e., how many months a customer has been with the company). The histogram below, faceted by contract type, shows that many customers are either very new (with a spike in the 0–5 month range) or have stayed for the maximum tenure of 72 months. This suggests the company has both a large group of recent sign-ups and a core of long-term, loyal customers, with fewer in the mid-range. Newer customers are at higher risk of churn, while long-tenured customers have demonstrated lasting retention.

```{r}
histo <- ggplot(telco, aes(x = tenure, fill = Churn)) +
  geom_histogram(binwidth = 5, color = "black", position = "stack") +
  facet_wrap(~ Contract) +
  scale_fill_manual(values = project_colors) +
  labs(
    title = "Distribution of Customer Tenure by Contract Type and Churn Status",
    x = "Tenure (months)",
    y = "Number of Customers",
    fill = "Churn"
  ) +
  theme_minimal()
histo
ggsave("histo.png", width = 8, height = 6, dpi = 300)
```
This visualization allows us to clearly observe how the likelihood of customer churn changes over time.

Churn rates are highest in the earliest months, particularly during the first 10–15 months of service. However, as tenure increases, churn steadily declines. The curve begins to flatten noticeably around the 20-month mark, suggesting that customers who stay beyond this point become significantly less likely to leave.

This creates a clear inflection point around 20 months, which can be strategically important for retention efforts. Rather than focusing only on long-tenured customers (e.g., 72 months, the max), telecom companies might benefit from targeting interventions and loyalty incentives around months 10–20, where the drop-off is steepest and loyalty is still being formed.

```{r}
churn_by_tenure <- telco %>%
  group_by(tenure) %>%
  summarize(
    churn_rate = mean(Churn == "Yes")
  )

inflection <- ggplot(churn_by_tenure, aes(x = tenure, y = churn_rate)) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE, color = "darkred") +
  labs(
    title = "Smoothed Churn Rate by Tenure",
    x = "Tenure (months)",
    y = "Churn Rate"
  )

plot <- ggplotly(inflection)

inflection

saveWidget(plot, "inflection.html", selfcontained = TRUE)


```

To explore how monthly charges differ across contract types, I used a boxplot to compare the distribution of charges for customers on Month-to-month, One-year, and Two-year plans. The results show that Month-to-month customers tend to pay slightly more, with a wider range of charges and more high-cost outliers. In contrast, customers on longer-term contracts have tighter, lower distributions, likely due to discounts or bundling associated with those plans—or potentially due to differences in the types of customers who opt for them.

```{r}
box <- ggplot(telco, aes(x = Contract, y = MonthlyCharges, fill = Contract)) +
  geom_boxplot() +
  scale_fill_manual(values = project_colors) +
  labs(
    title = "Monthly Charges by Contract Type",
    x = "Contract Type",
    y = "Monthly Charges (USD)"
  ) +
  theme_minimal()
box
ggsave("box.png", width = 8, height = 6, dpi = 300)

```
Rather than speculating further, I decided to run a multiple linear regression to better understand which factors influence monthly charges. The regression results indicate that Internet service type, contract length, and optional services (such as StreamingTV and TechSupport) significantly influence monthly charges. Specifically:
Fiber optic internet and month-to-month contracts are strong predictors of higher charges.
Longer contracts and no internet service are associated with lower charges.
Add-on services also drive costs upward, reflecting more engaged or high-value customers.
These insights align with customer segmentation patterns and may inform targeted pricing strategies or personalized plan recommendations.
To visualize the regression output, I provide a bar chart of coefficient estimates.


```{r}
reg_data <- telco %>%
  select(-TotalCharges) %>%
  na.omit()  

reg_data <- reg_data %>%
  mutate(across(where(is.character), as.factor))

model <- lm(MonthlyCharges ~ ., data = reg_data)

summary(model)

```


```{r}
coeffs <- broom::tidy(model) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(estimate))

coef <- ggplot(coeffs, aes(x = reorder(term, estimate), y = estimate)) +
geom_col(fill = project_colors["Yes"]) +
  coord_flip() +
  labs(
    title = "Regression Coefficients: Impact on Monthly Charges",
    x = "Predictor",
    y = "Estimated Impact ($)"
  ) +
  theme_minimal()

coef
ggsave("coef.png", width = 8, height = 6, dpi = 300)


```
To explore the interaction between contract type and internet service on churn behavior, I created a heatmap showing the churn rate across all combinations of these two categorical variables. Each tile represents a unique pairing, with color intensity indicating churn percentage—lighter shades represent lower churn, and deeper reds indicate higher churn.

Month-to-month + Fiber optic customers show the highest churn rate, represented by the most intense red tile. This combination likely reflects customers who face high costs without the commitment of a contract, making them more likely to cancel.

In contrast, Two-year contracts, regardless of internet service, consistently exhibit low churn (often below 10%), shown by the lightest tiles. This reinforces earlier findings about long-term contracts being associated with stronger retention.

DSL users on month-to-month plans have a moderate churn rate—higher than two-year DSL contracts, but not as high as fiber. This could reflect cost differences between DSL and fiber internet.

Interestingly, customers with no internet service on month-to-month plans (likely phone-only users) show relatively low churn, potentially due to lower billing amounts or fewer alternatives.


```{r}
churn_rate <- telco %>%
  group_by(Contract, InternetService) %>%
  summarize(churn_pct = mean(Churn == "Yes") * 100)

ggplot(churn_rate, aes(x = Contract, y = InternetService, fill = churn_pct)) +
  geom_tile(color="white") +
  scale_fill_gradient(low="lightyellow", high="red", name="Churn Rate (%)") +
  labs(title="Churn Rate Heatmap: Contract vs Internet Service", 
       x="Contract Type", y="Internet Service") +
  theme_minimal()

```
To explore how churn relates to both tenure and monthly charges, I created a scatter plot where each point represents a customer, colored by churn status.

The visualization reveals that churned customers are heavily concentrated at low tenure levels, supporting earlier findings that many cancellations occur early in the customer lifecycle, often among those on month-to-month contracts. In contrast, churn is rare among long-tenured customers, indicating stronger loyalty or satisfaction over time.

There is also a noticeable cluster of churned customers in the upper-left region of the plot—those with high monthly charges and low tenure—suggesting that new customers with higher bills are more likely to cancel early.

This pattern reinforces the inverse relationship between tenure and churn, and suggests that monthly charges may also contribute to early cancellations. These insights highlight potential opportunities for improving early retention through pricing adjustments, onboarding enhancements, or targeted support for high-risk customer segments.

```{r}
telco$Churn <- factor(trimws(telco$Churn), levels = c("No", "Yes"))

churn <- ggplot(telco, aes(x = tenure, y = MonthlyCharges, color = Churn)) +
  geom_point(alpha = 0.6) +
  labs(title = "Monthly Charges vs Tenure, by Churn Status",
       x = "Tenure (months)", y = "Monthly Charges (USD)") +
  scale_color_manual(values = project_colors[c("No", "Yes")]) +
  theme_minimal()
churn 
ggsave("churn.png", width = 8, height = 6, dpi = 300)


```





