---
title: "MidTerm_Telecom"
author: "Emma Horton"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

MidTerm Telecom Customer Churn Analysis
Emma Horton – Data Visualization Midterm Project

Libraries Used: 
```{r libraries, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(plotly)
library(lsr)
library(broom)

```

Data Loading
```{r}
telco <- read.csv("Telco-Customer-Churn.csv", na.strings = c(""))

#str(telco)
#summary(telco)
#head(telco, 5)
```
Data Cleaning

```{r}
# SeniorCitizen to factor
telco$SeniorCitizen <- factor(telco$SeniorCitizen, levels=c(0,1), labels=c("No", "Yes"))

# Character columns to factors
cat_cols <- c("gender","Partner","Dependents","PhoneService",
              "MultipleLines","InternetService","OnlineSecurity","OnlineBackup",
              "DeviceProtection","TechSupport","StreamingTV","StreamingMovies",
              "Contract","PaperlessBilling","PaymentMethod","Churn")

telco[cat_cols] <- lapply(telco[cat_cols], factor)

# Fix TotalCharges: convert to numeric
telco$TotalCharges <- as.numeric(as.character(telco$TotalCharges))

# Replace "No internet service" and "No phone service" with "No", replace "No phone service" with "No"
no_internet_cols <- c("OnlineSecurity","OnlineBackup","DeviceProtection",
                      "TechSupport","StreamingTV","StreamingMovies")
for(col in no_internet_cols) {
    telco[[col]] <- factor(ifelse(telco[[col]]=="No internet service", "No", 
                                  as.character(telco[[col]])))
}
telco$MultipleLines <- factor(ifelse(telco$MultipleLines=="No phone service", "No", 
                                     as.character(telco$MultipleLines)))
# Drop customerID
telco$customerID <- NULL

# Verify results
#sum(is.na(telco$TotalCharges))  
#levels(telco$MultipleLines)        
#levels(telco$OnlineSecurity)       

```
Set Color Pallet
```{r}
project_colors <- c(
  "Yes" = "#E74C3C",    
  "No" = "#3498DB",         
  "Female" = "#9B59B6",    
  "Male" = "#1ABC9C",     
  "Month-to-month" = "#F39C12",  
  "One year" = "#2ECC71",       
  "Two year" = "#34495E",       
  "DSL" = "#16A085",        
  "Fiber optic" = "#D35400",
  "No" = "#95A5A6"          
)

```

Exploratory Analysis
Distribution of a single categorical variable
1. Categorical Variable Distribution – Contract Type vs Churn (Bar Chart)
A key categorical variable is the Contract type. I suspect contract length has an effect on churn. I will plot the count of customers by contract type, and distinguish churn vs not churn with color.

```{r}
contract_churn_data <- telco %>%
  group_by(Contract, Churn) %>%
  summarise(count = n(), .groups = 'drop')

plot_ly(contract_churn_data,
        x = ~Contract,
        y = ~count,
        color = ~Churn,
        colors = project_colors,
        type = "bar") %>%
  layout(
    barmode = "group",
    title = list(text = "Customer Count by Contract Type and Churn Status", x = 0.5),
    xaxis = list(title = "Contract Type"),
    yaxis = list(title = "Number of Customers"),
    legend = list(title = list(text = "Churn"))
  )

```
This is a strong indicator that contract length is related to churn. I will perform a deeper dive into the relationship

```{r}
# Cross-tab Churn by Contract
table(telco$Churn, telco$Contract)

# Chi-square
chisq_test <- chisq.test(table(telco$Churn, telco$Contract))
chisq_test

# mosaic 
mosaicplot(table(telco$Contract, telco$Churn),
           color = c(project_colors["No"], project_colors["Yes"]),
           main = "Churn vs Contract Type",
           xlab = "Contract Type", ylab = "Churn")

```

Distribution of a single quantitative variable
2. Numerical Variable Distribution – Tenure (Histogram)
I want to explore the distribution of tenure (how long customers have been with the company, in months). A histogram will show the frequency of customers by their tenure.
```{r}
ggplot(telco, aes(x = tenure)) +
  geom_histogram(binwidth = 5, fill=project_colors["No"], color="black") +
  labs(title="Distribution of Customer Tenure",
       x="Tenure (months)", y="Number of Customers") +
  theme_minimal()
```
I want to explore the representation of churn status in regards to tenure, faceting on contract type.

```{r}
ggplot(telco, aes(x = tenure, fill = Churn)) +
  geom_histogram(binwidth = 5, color = "black", position = "stack") +
  facet_wrap(~ Contract) +
  scale_fill_manual(values = project_colors) +
  labs(
    title = "Distribution of Customer Tenure by Contract Type and Churn Status",
    x = "Tenure (months)",
    y = "Number of Customers",
    fill = "Churn"
  ) +
  theme_minimal()
```

Distribution of two categorical variables
3. Two Categorical Variable Distribution – Payment Method vs. Churn (Grouped Bar)
I want to explore the relationship between payment method (bank transfer, credit card, electronic / mailed check) and churn. A grouped bar will display the relationship between payment method and churn.


```{r}
payment_churn_data <- telco %>%
  group_by(PaymentMethod, Churn) %>%
  summarise(count = n(), .groups = "drop")

# Remove "(automatic)" from PaymentMethod labels
telco$PaymentMethod <- gsub("\\s*\\(automatic\\)", "", telco$PaymentMethod)


plot_ly(payment_churn_data,
        x = ~PaymentMethod,
        y = ~count,
        color = ~Churn,
        colors = project_colors,
        type = "bar") %>%
  layout(
    barmode = "group",
    title = list(text = "Customer Churn by Payment Method", x = 0.5),
    xaxis = list(title = "Payment Method", tickangle = 30),
    yaxis = list(title = "Number of Customers"),
    legend = list(title = list(text = "Churn"))
  )


```

I want to explore how contract type will impact the distribution
```{r}
ggplot(telco, aes(x = PaymentMethod, fill = Churn)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ Contract) +
  scale_fill_manual(values = project_colors) +
  labs(
    title = "Churn Distribution by Payment Method and Contract Type",
    x = "Payment Method",
    y = "Customer Count",
    fill = "Churn"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))


```
4. Quantitative Variable Across Categories – Monthly Charges by Contract Type (Boxplot)
I want explore how monthly charges differ by contract type.
```{r}
ggplot(telco, aes(x = Contract, y = MonthlyCharges, fill = Contract)) +
  geom_boxplot() +
  scale_fill_manual(values = project_colors) +
  labs(
    title = "Monthly Charges by Contract Type",
    x = "Contract Type",
    y = "Monthly Charges (USD)"
  ) +
  theme_minimal()

```
Instead of mucking around to figure out the reason why I suspect a quick regression will give me more answers in less time.

Quick regression
```{r}
reg_data <- telco %>%
  select(-TotalCharges) %>%
  na.omit()  

reg_data <- reg_data %>%
  mutate(across(where(is.character), as.factor))

model <- lm(MonthlyCharges ~ ., data = reg_data)

summary(model)

```

To display the regression, I will display a coefficient Plot Bar Chart of Estimates
```{r}
coeffs <- broom::tidy(model) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(estimate))

ggplot(coeffs, aes(x = reorder(term, estimate), y = estimate)) +
geom_col(fill = project_colors["Yes"]) +
  coord_flip() +
  labs(
    title = "Regression Coefficients: Impact on Monthly Charges",
    x = "Predictor",
    y = "Estimated Impact ($)"
  ) +
  theme_minimal()

```

```{r}
ggplot(telco, aes(x = InternetService, y = MonthlyCharges, fill = InternetService)) +
  geom_boxplot() +
  scale_fill_manual(values = c(
    "DSL" = "#16A085",
    "Fiber optic" = "#D35400",
    "No" = "#95A5A6"
  )) +
  labs(
    title = "Monthly Charges by Internet Service Type",
    x = "Internet Service",
    y = "Monthly Charges (USD)"
  ) +
  theme_minimal()

```
Monthly Charges by Streaming TV Usage
```{r}
ggplot(telco, aes(x = StreamingTV, y = MonthlyCharges, fill = StreamingTV)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Yes" = "#E74C3C", "No" = "#3498DB")) +
  labs(
    title = "Monthly Charges by Streaming TV Subscription",
    x = "Streaming TV",
    y = "Monthly Charges (USD)"
  ) +
  theme_minimal()

```
Monthly Charges by Tech Support
```{r}
ggplot(telco, aes(x = TechSupport, y = MonthlyCharges, fill = TechSupport)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Yes" = "#E74C3C", "No" = "#3498DB")) +
  labs(
    title = "Monthly Charges by Tech Support Access",
    x = "Tech Support",
    y = "Monthly Charges (USD)"
  ) +
  theme_minimal()

```
Monthly Charges by Paperless Billing
```{r}
ggplot(telco, aes(x = PaperlessBilling, y = MonthlyCharges, fill = PaperlessBilling)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Yes" = "#E74C3C", "No" = "#3498DB")) +
  labs(
    title = "Monthly Charges by Paperless Billing Status",
    x = "Paperless Billing",
    y = "Monthly Charges (USD)"
  ) +
  theme_minimal()

```
5. Relationship between two quantitative variables
Relationship Between Two Variables – Tenure vs Monthly Charges (Scatter Plot)
Now I explore how a numeric variable relates to churn. Plotting MonthlyCharges against tenure, and using color to show churn status, can reveal if churners cluster in some region.

```{r}
telco$Churn <- factor(trimws(telco$Churn), levels = c("No", "Yes"))

ggplot(telco, aes(x = tenure, y = MonthlyCharges, color = Churn)) +
  geom_point(alpha = 0.6) +
  labs(title = "Monthly Charges vs Tenure, by Churn Status",
       x = "Tenure (months)", y = "Monthly Charges (USD)") +
  scale_color_manual(values = project_colors[c("No", "Yes")]) +
  theme_minimal()

```

Create a chart not directly discussed in class (e.g. Heatmap, Radar, Contour, Sunburst)
6. Heatmap – Churn Rate by Contract and Internet Service
Finally, I wanted to create a heatmap to visualize churn rates across two categorical dimensions: Contract type and InternetService type. This will highlight combinations (e.g. Fiber optic + month-to-month) that have especially high churn.

```{r}
churn_rate <- telco %>%
  group_by(Contract, InternetService) %>%
  summarize(churn_pct = mean(Churn == "Yes") * 100)

ggplot(churn_rate, aes(x = Contract, y = InternetService, fill = churn_pct)) +
  geom_tile(color="white") +
  scale_fill_gradient(low="lightyellow", high="red", name="Churn Rate (%)") +
  labs(title="Churn Rate Heatmap: Contract vs Internet Service", 
       x="Contract Type", y="Internet Service") +
  theme_minimal()

```
7. Bubble Plot – Customer Charges Over Time by Contract Type
 I wanted to highlight how customers on month-to-month contracts often accumulate high charges earlier, while those on longer contracts tend to accrue charges more gradually.
 
 
 
```{r, echo=FALSE, message=FALSE, warning=FALSE}
telco$TenureBin <- cut(telco$tenure,
                       breaks = seq(0, 72, by = 6),
                       include.lowest = TRUE,
                       labels = c("0–6", "7–12", "13–18", "19–24", "25–30", "31–36",
                                  "37–42", "43–48", "49–54", "55–60", "61–66", "67–72"))

telco$TotalCharges <- as.numeric(as.character(telco$TotalCharges))

telco_clean <- telco %>% filter(!is.na(TotalCharges))


plot_ly(
  data = telco_clean,
  x = ~MonthlyCharges,
  y = ~TotalCharges,
  size = ~tenure,
  color = ~Contract,
  frame = ~TenureBin,
  text = ~paste("Contract:", Contract,
                "<br>Monthly Charges:", round(MonthlyCharges, 2),
                "<br>Total Charges:", round(TotalCharges, 2),
                "<br>Tenure:", tenure),
  colors = c(
    "Month-to-month" = "#F39C12",
    "One year" = "#2ECC71",
    "Two year" = "#34495E"
  ),
  type = 'scatter',
  mode = 'markers',
  marker = list(opacity = 0.7, sizemode = "area", line = list(width = 1, color = "#FFFFFF"))
) %>%
  layout(
    title = list(text = "Customer Charges Over Time by Contract Type", x = 0.5),
    xaxis = list(title = "Monthly Charges"),
    yaxis = list(title = "Total Charges"),
    showlegend = TRUE
  )


```

